---
title: "Data Visualization"
description: |
  Mono Lake water levels and management interventions.
author:
  - name: Lucas Boyd
date: 2022-01-17
output:
  distill::distill_article:
    code_folding: code
    self_contained: false
---

### Setup 
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# attach packages
library(tidyverse)
library(janitor)
library(here)
library(scales)
```
### Data wrangling
```{r}
# reading in the mono lake data
mono <- read_csv(here("data", "mono_lake_annual_levels.csv")) %>% 
# turning a row into column names
  row_to_names(row_number = 5) 
# creating another dataset with cleaner names to work with and selecting certain variables
mono_clean <- mono %>% 
  clean_names() %>% 
# changing the class of variables from characters to numerics
  mutate(level = as.numeric(lake_level_feet_above_sea_level)) %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(stable_level = as.numeric(stable_level)) %>% 
  select(year, level, stable_level)

# creating a subset to add geom points
 mono_point <- mono_clean %>% 
  filter(year %in% c(1850, 1941, 2017, 1978))
  
```
### Visualization
```{r}
ggplot(data = mono_clean, aes(x = year, y = level)) +
  geom_hline(yintercept = 6392, # adding a line for the water management level
             color = "cyan4", 
             linetype = "dashed", size = 1) +
  annotate("text", x = 1867, y = 6395, # adding a label
            label = "Stability Goal", 
            color = "cyan4") +
  theme_classic(12) + 
 scale_y_continuous(limits = c(6345,6430)) +
  labs(y = "Water Level (feet above sea level)", x = "Year") +
  geom_rect(aes(ymin=6360, ymax = 6377, xmin = -Inf, xmax = Inf, # adding shaded regions to show adverse affects of lowering water levels
                fill = "Land Bridges Form")) +
  geom_rect(aes(ymin=-Inf, ymax = 6350, xmin = -Inf, xmax = Inf, 
                fill = "Acute Food Shortage")) +
   geom_rect(aes(ymin=6350, ymax = 6360, xmin = -Inf, xmax = Inf, 
                 fill = "Food Affected")) +
  scale_fill_manual(values = c("firebrick1", "darkorange1", "gold"), name = element_blank()) + # manually changing colors
    geom_line(size = 1.2) +
  scale_x_continuous(limits = c(1850,2017), # changing number of ticks on x axis
                     breaks = pretty(x = mono_clean$year, n = 10)) +
    annotate("text", x = 1868, y = 6368, # adding labels for shaded areas
              label = "Land Bridges Form") +
  annotate("text", x = 1870, y = 6355, 
            label = "Food Supply Affected") +
  annotate("text", x = 1870, y = 6345, 
            label = "Acute Food Shortage") +
  theme(legend.position = "none") + # removing legend
  geom_point(data = mono_point, aes(x = year, y = level), 
             size = 3, shape = 21, color = "grey27", fill = "black") + # adding points of interest
  annotate("text", x = 1971, y = 6421, 
           label = "1941: Water Diverted to Los Angeles", 
           color = "gray27") + #labeling those points
    annotate("text", x = 2017, y = 6385, 
             label = "2017", 
             color = "gray27") +
    annotate("text", x = 1850, y = 6410, 
             label = "1850", 
             color = "gray27") +
  annotate("text", x = 1978, y = 6369, 
           label = "1978: MLC Founded", 
           color = "gray27")

```

**Fig. 1** shows the water level (in feet above sea level) for Mono Lake from 1850 until 2017 (indicated by the black line). In 1941, the Los Angeles Department of Water & Power began diverting water from Mono Lake's tributary streams to meet the water demands of the city. In 1978, the Mono Lake Committee was founded and set a goal to restore the lake to previous levels. The dashed line represents MLC's target of stabilizing the water level at 6392 feet. Critical levels for aquatic birds are indicated by the shaded regions. At a water level of 6377 feet, land bridges form and expose nesting sites to predators. At 6360 feet, high salinity decreases food availability for aquatic birds. At 6350 feet, the impacts of the food shortage become acute. 

