---
title: "Binomial Logistic Regression"
description: |
  Palmetto species analysis and model fit. 
author:
  - name: Lucas Boyd
    url: {}
date: 2022-02-18
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# attach packages
library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(AICcmodavg)
library(caret)
library(kableExtra)
library(equatiomatic)
```

### Overview

The following report examines growth data of two dominant palmetto
species (*Serenoa repens* and *Sabal etonia*) in South Florida to
produce a binomial logistic regression model that classifies plant
species based on several physical features of the plant. First,
differences between the species are visualized to determine which
variables will be useful in classifying species. Then, the model is
evaluated using 10-fold cross validation and AIC. Finally, the model is
applied to the data to determine its efficacy in predicting plant
species.

**Data Citation:** Data source: Abrahamson, W.G. 2019. Survival, growth
and biomass estimates of two dominant palmetto species of south-central
Florida from 1981 - 2017, ongoing at 5-year intervals ver 1.
Environmental Data Initiative.
<https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5>

### Reading in and cleaning up the data

```{r}
palmetto <- read_csv(here("data", "palmetto.csv")) %>% 
  select(height, length, width, green_lvs, species) %>% # selecting for variables
# 1 = Serenoa repens, 2 = Sabal etonia
  drop_na() %>% # dropping NA values
  mutate(species_name = case_when(
    species == 1 ~ "Serenoa repens", # new column for species name
    species == 2 ~ "Sabal etonia"
  )) %>% 
  mutate(species_name = as_factor(species_name)) # changing species name to a factor for use in blr
```

**Variables of interest:**
-   Number of green leaves - green_lvs

-   Height of plant (cm) - height

-   Width of canopy (cm) - width

-   Length of canopy (cm) - length

### Visualizing differences between species

```{r}
# ggplots to visualize the differences in each variable between species
# effectplot could also be useful in these visualizations for blr 

p1 <- ggplot(data = palmetto, aes(x = species_name, y = green_lvs, fill = species_name)) + 
  geom_violin(scale = "count", color = "black") + # geom violin to visualize the large volume of data
geom_boxplot(color = "black", fill = NA, width = 0.1, outlier.color = NA) + # adding summary median and quartiles to figure
  stat_summary(fun=mean, # adding the mean to the figure 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black") +
  scale_fill_manual(values = c("orange4", "darkolivegreen")) + # customizing colors
  theme_minimal(13) + # changing theme and font size
  theme(legend.position = "top",
        axis.text.x = element_blank(),
        legend.title = element_blank()) +
  labs(x = element_blank(), y = "Number of green leaves") # adding labels

# repeated above code for each variable in subsequent 3 plots, inserting different variables

p2 <- ggplot(data = palmetto, (aes(x = species_name, y = height, fill = species_name))) +
    geom_violin(scale = "count", color = "black") +
geom_boxplot(color = "black", fill = NA, width = 0.1, outlier.color = NA) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black") +
  scale_fill_manual(values = c("orange4", "darkolivegreen")) +
  theme_minimal(13) +
  theme(legend.position = "none",
        axis.text.x = element_blank()) +
  labs(x = element_blank(), y = "Plant Height (cm)") 

p3 <- ggplot(data = palmetto, (aes(x = species_name, y = width, fill = species_name))) +
    geom_violin(scale = "count", color = "black") +
geom_boxplot(color = "black", fill = NA, width = 0.1, outlier.color = NA) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black") +
  scale_fill_manual(values = c("orange4", "darkolivegreen")) +
  theme_minimal(13) +
  theme(legend.position = "none",
        axis.text.x = element_blank()) +
  labs(x = element_blank(), y = "Canopy Width (cm)") 

p4 <- ggplot(data = palmetto, (aes(x = species_name, y = length, fill = species_name))) +
    geom_violin(scale = "count", color = "black") +
geom_boxplot(color = "black", fill = NA, width = 0.1, outlier.color = NA) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=4, 
               color="black", 
               fill="black") +
  scale_fill_manual(values = c("orange4", "darkolivegreen")) +
  theme_minimal(13) +
  theme(legend.position = "none",
        axis.text.x = element_blank()) +
  labs(x = element_blank(), y = "Canopy Length (cm)") 
```

```{r}
(p1 | p2)
(p3 | p4)
```

**Fig. 1** compares two palmetto species to determine feasibility of
each of four variables in predicting whether a plant is *Serenoa repens*
(green) or *Sabal etonia* (orange). Variables examined in the figure
are: number of green leaves on the plant (top left), height of plant
(cm) (top right), canopy width (cm) (bottom left) and canopy length (cm)
(bottom right). Width of the shape in each figure represents the
quantity of observations at that particular value. Box endpoints
indicate the 25^th^ and 75^th^ percentile values; the black line and
black point within the box indicate the median and mean value for each
species, respectively.

**Conclusion:** Based on this visualization, it appears that number of
green leaves and canopy length will be useful in correctly classifying
species, since there is a noticeable difference in the mean, median and
overall spread of the data. Plant height and canopy width differ only
very slightly between the two species, and thus will be less useful in
classifying the species of plant.

### Binomial logistic regression

Running BLR for two models:

-   **Model 1:** Plant species as a function of plant height, number of
    green leaves, canopy width and canopy length

-   **Model 2:** Plant species as a function of plant height, number of
    green leaves, and canopy width

```{r}
# storing functions of combinations of variables for use in regression
f1 <- species_name ~ height + length + width + green_lvs
f2 <- species_name ~ height + width + green_lvs

# running and storing the regressions for each formula
blr1 <- glm(formula = f1,
            data = palmetto,
            family = "binomial")

blr2 <- glm(formula = f2,
            data = palmetto,
            family = "binomial")

# taking a look at the outputs of the regression with summary()
# reference (0) = Sabal etonia
```

### Model Selection

#### AIC

```{r}
aic <- aictab(list(blr1, blr2))
# model1 = 5194
# model2 = 5987
```

**Model 1:** `r round(aic$AICc[1], 1)`

**Model 2:** `r round(aic$AICc[2], 1)`

The difference in AICc is very large (`r round(aic$Delta_AICc[2], 1)`), thus model
1 has significantly better fit than model 2.

#### 10-fold cross validation

```{r}
set.seed(24) # setting seed for reproducibility

tr_ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10) # storing train control, specifying 10 folds

# train the model
model1 <- train(f1, data = palmetto,
                method = "glm",
                family = "binomial",
                trControl = tr_ctrl)
# accuracy = 0.917

# train the other model
model2 <- train(f2, data = palmetto,
                method = "glm",
                family = "binomial",
                trControl = tr_ctrl)
# accuracy = 0.899
```

**Model assessment results:** Model 1, which classifies palmetto species
as a function of plant height, canopy width, canopy length and number of
green leaves, performed better during 10-fold cross validation (accuracy
= 91.7%, AICc = `r round(aic$AICc[1], 1)`) than model 2 (accuracy = 89.9%, AICc =
`r round(aic$AICc[2], 1)`). Thus, based on the prediction accuracy results and AIC
values, model 1 is preferred.

### Regression results

**Table 1** shows the results of binomial logistic regression for model
1. Coefficients show the change in log odds of the resulting species
being *Sabal etonia* (reference) as opposed to *Serenoa repens*.

```{r}
# storing the results of regression as a data frame
blr1_tidy <- tidy(blr1) %>%  
  mutate(p.value = case_when( # finalizing p.values for table
    p.value <0.001 ~ "<0.001"
  )) %>% 
  mutate(term = case_when( # finalizing variable names for table
         term == "(Intercept)" ~ "Intercept",
         term == "height" ~ "Height",
         term == "length" ~ "Length",
         term == "width" ~ "Width",
         term == "green_lvs" ~ "# of green leaves")) %>% 
  select(-statistic)

# finalizing table
kable(blr1_tidy, 
      col.names = c("Term", "Coefficient", "Standard Error", "P Value"),
      digits = 3) %>% 
  kable_minimal(full_width = FALSE)
  
```

**Final Model Equation**:
`r equatiomatic::extract_eq(blr1, wrap = TRUE, use_coefs = TRUE)`

### Model success rate

**Table 1** shows the success rate of the chosen model in classifying
each plant species when trained on and applied to the whole data set.

```{r}
# storing a function for prediction accuracy
prediction_accuracy <- function(x, y) {
  accurate <- ifelse(x == y, 1, 0)
  return(accurate)
}

# creating a new dataframe with predictions based on our model
# could also use augment() here: palmetto_augment <- blr1 %>% augment(type.predict = "response") 
palmetto_success <- palmetto %>% 
  mutate(chances_sabal = predict(blr1, # adding a new column for the probability that the species is sabal, given our model
                   palmetto,
                   type = "response")) %>% 
  mutate(predicted = ifelse(chances_sabal >= .50, "Sabal etonia", "Serenoa repens")) %>%  
  # adding a new column that predicts the species based on the probability, using 50% as the threshold
  mutate(accurate = prediction_accuracy(species_name, predicted))

# creating a final table for model prediction accuracy for each plant

accuracy_table <- palmetto_success %>% 
  group_by(species_name) %>% # grouping by species
  summarize(observations = n(),
    total_accurate = sum(accurate), # finding total accurate predictions
            total_inaccurate = n() - sum(accurate), # total inaccurate predictions
            accuracy = (mean(accurate))*100,) # accuracy %

# finalized table with kable
kable(accuracy_table, digits = 1, col.names = c("Palmetto Species", "Total Observations", "Accurate", "Inaccurate", "% Accurate ")) %>% 
  kable_minimal(full_width = FALSE)


```

**Conclusion:** Model 1 (species as a function of plant height, canopy
width, canopy length, and number of green leaves) is highly effective in
classifying a plant as either *Sabal etonia*
(`r round(accuracy_table$accuracy[2], 1)`%, n =
`r accuracy_table$observations[2]`) or Serenoa repens
(`r round(accuracy_table$accuracy[1], 1)`%, n =
`r accuracy_table$observations[1]`).
